---
title: "Explicit Timecourse"
output: html_notebook
editor_options:
  markdown:
    wrap: 72
---

# Overview

When participants experience a cursor rotation while making a reach to a
target, they must compensate by deviating their hand movement in order
to get a cursor straight to its target. While the strategies participants 
implement to achieve this are thought to emerge gradually, recent data
in our lab has shown that strategies might instead develop as a stepwise
process (D'Amario et al.)

In this experiment, we aimed to identify the timecourse of explicit
adaptation. Participants learned to counter a rotation and were
instructed to report their aiming strategy by changing the direction of
an arrow.

Rotation sizes introduced were 20, 30, 40, 50, and 60 degrees

<br>

# Data

We will be exploring: <br>1. The proportion of learners within each
perturbation condition. <br>2. The proportion of aiming strategies as a
function of rotation size. <br>3.The Emergence of distinct patterns across strategy-users.

# Setup

Get necessary packages:

##renv
```{r}

if (!requireNamespace("renv", quietly = TRUE)) install.packages("renv")
library(renv)

renv::restore(prompt = FALSE)

```



```{r}
library('remotes')
ip <- installed.packages()
if ('Reach' %in% ip[,'Package']) {
  if (ip[which(ip[,'Package'] == 'Reach'),'Version'] < "2025.02.16") {
    remotes::install_github('thartbm/Reach')
  }
} else {
  remotes::install_github('thartbm/Reach')
}
```



Source project code:

```{r}
# load sources for this project
source("R/data.R") 
source("R/newcode.R") 
source("R/AimDevPlot.R") 
source("R/Clustering/learningAlg.R") 
source("R/Clustering/KmeansXG.R") 
source("R/Clustering/clusterBehaviours.R")
source("R/Reach Data/ReachDevPlot.R")
source("R/Reach Data/nocursorReaches.R")
source("R/newstats.R") 
```

Download Files
```{r}
getData()
```

## Analysis

### Learning

Did all participants learn to counter the perturbation at least 50% of
the time?

```{r}
LearnerCSV()
```

This is determined by ensuring the baseline-corrected reachadeviation_deg median of the
final 16 rotated trials is greater than 50% of the rotation magnitude
received.

Then we can filter out non-learners


We can plot participant reach deviations to show the magnitude of
learning per rotation group

```{r}
plotREACH()

```

### Aiming

[Aiming analysis script](newcode.R)

Did all learners have a strategy (ie, deviate their aim)?

```{r}
strategySummary()
```

We can use a 95% CI approach to determine strategy-users. A strategy
will be defined to: 1) Within the last 8 trials of the rotation phase,
each CI bound must be above 0 degrees 2) The lower bound must be above 5
degrees to account for motor noise

For further analysis, we can create a csv file for only strategy users.

```{r}
Strategyfile()
strat_data <- read.csv("data/strategy_only_participants.csv")
```

#### Statistics

[Aiming stats script](newstats.R)

**1) Does final aligned and first rotated aims differ from \~0**

```{r}
zeroT()
```

Here we took the mean final 16 aligned and mean First 16 rotated aim deviations per
participant and ran one sided T-tests per rotation group. Aligned aim deviations to 0 is significant (p<0.001), indicating aim biases in the clockwise direction. First rotated aims to 0 is significant (p < 0.001).

**2) Is likelihood of strategy development predicted by rotation size?**
- Logistic regression analysis to predict probability of developing a strategy.
```{r}
logAnalysis()
```

The odds of using a strategy are above 0 in the 20 degree group, however the odds of using a strategy is higher in all other rotations as compared to the 20 degree group.

**3)Does aim magnitude vary across rotation sizes?**

```{r}
aimAOV()
```

Taking the mean from the last 16 trials, One way anova shows an effect
of rotation size (p<0.0001). Post-hoc tukey shows that the 60 degree group produces
a higher aim strategy than all other groups. 50 degree is higher than 30 and 20 group. 40 degree is higher than the 20 group. 

**4) Does aim variance vary across rotation sizes?**
<br> - One way ANOVA (IV: Rotation, DV: final SD of aim) with Tukey test

```{r}
aimvarAOV()
```

Taking the sd from the last 16 trials, one way anova shows an effect of
rotation size (p=0.0005). Post-hoc tukey shows that participants in the 60°
rotation group show significantly greater variability in their final
performance than participants in the 20°, 30°, and 40° groups.

**5) Is there an effect of rotation size for the timepoint of strategy-onset?**
<br> - One-way ANOVA (IV: rotation, DV: trial)
```{r}
startTrialAOV()
```

#### Visualization

**Aiming Behaviour**
```{r}
plotAIM2()
```


### Are there 'patterns' in strategy development among participants?

Inspired by D'Amario et al 2024 which first showed aiming patterns different from 
an inverted exponential decay function, we noticed that not all participants develop
strategies solely as a step or as an exponential process. We note different patterns as well.
Tsay et al. 2024 initially suggested distinct classifications of
strategies. We can therefore conduct a kMeans cluster analysis to identify patterns
in our own data to determine if phenotypes other than gradual and stepwise exist.

**1) Predicting individual participant learning phases**

- First, we need to define a 'learning phase' finding the point of strategy onset to  when a strategy becomes stable,       using a rolling window  analysis approach, 
- Trial start: when a participants aim deviation exceeds the motor noise threshold of 5 degrees.
- Trial end: when the SD of a participants aim deviation closely matches the SD of the final 16 aiming trials and 
  remains within an SD of 1.9.

```{r}
xgRun()
```

**We can also visually assign these timepoints and compare these classifcations to the XgBoost model's predictions:**

```{r}
plotStart()
```
```{r}
plotEnd()
```



**2) Run a PCA participant features to capture aim heterogeneity**

FEATURES:
<br> 1) SD of learning phase
<br> 2) Length of learning phase
<br> 3) Absolute trial to trial changes
<br> 4) Number of clockwise (negative) aims
<br> 5) largest_jump_frac: Measures distribution of changes - Low → changes are more distributed (gradual or multiple smaller steps)
<br> 6) jump_ratio: Highest aim jump / mean of jump magnitudes - How extreme the largest jump is compared to typical jumps. Measures contrast between big vs small steps
<br> 7) Smoothness in aiming - normalized sum of absolute trial-to-trial changes, providing a measure of local variability in the learning trajectory
<br> 7) linearity - using the R² of a linear regression of aim over trials, capturing the extent to which behavior followed a consistent linear trend
<br> 9) max jump norm - Largest jump relative to total learning range. (High → one jump accounts for a large portion of total learning)
<br> 10) number of aim changes from clockwise to counterclockwise directions



Run PCA to determine most highest feature loadings:
```{r}
extract <- getFeaturesFromModel() 
  
model_df <- xgRun()
strategy_data <- read.csv("data/strategy_only_participants.csv")

features_df <- extract$features_df        
k_input     <- extract$kmeans_input     
k_scaled <- scale(k_input)

# -------------------------------
# PCA 
pca <- prcomp(k_scaled, center = TRUE, scale. = FALSE)
```
HIGHEST FEATURE LOADINGS within PC1 AND PC2:
<br> - PC1 – length duration & sign flips (negative so smaller = higher)
<br> - PC2 - Learning variance (sd and trial-trial changes are negative so smaller =higher)
<br> - PC3 - Jump size & Linearity (negative)

For plotting we can pca$x*-1 to visualize patterns easier
We can see which centres closely match participants' behavioural features by plotting the informative components in a 2D figure. <br> 

```{r}
plotComponents()
```
Thus, we have found three clusters or 'learning phenotypes that describe strategy heterogeneity. 


**2) Determine which components are most informative**
We can create a plot that assesses cumulative variance explained across PC components. Here, we can calculate the variance of each component and divide by the variance sum. Looks like the first 3 components make the largest contribution to the total variance (~83%) with steeper slopes.
```{r}
plotScree()
```


**3) Run an unsupervised kMeans clustering analysis on three most informative components**

Compare clusters to visually predicted clusters
```{r}
confMatrix()
```


**Is cluster predicted by rotation size?**
<br> - We can run a fisher's exact test with a monte carlo simulation to account for small cell count and large table frequency.

```{r}
rotationEffect() 
```



#### What do group averages per cluster look like? Do we still find an inverted exponential decay function?
```{r}
normalizeClusters()
```

To determine whether asymptotic magnitude trial of reached asymptotic, variability of slope differ across cluster groups:
```{r}
clusterMetrics()
```



### No cursor reaches

After 120 rotated trials, participants undergo 24 trials of no-cursor reaches. We can use these trials to assess aftereffects and whether it changes across strategy/non-strategy-users and strategy phenotype.

####1) Do aftereffects differ across strategy levels including non learners?


First lets visualize washout in non learners :

```{r}
nonCSV()
```


Visualize washout in non strategy learners : 

Visualize washout in the exploratory strategy group :

Visualize washout in the stepwise strategy group :

Visualize washout in the gradual strategy group :

Run 4x5 mixed anova (rotation as within subs, strategy group as between) - exclude non Learners 
```{r}
fiveANOVA()
```
Run baysian 4x5 anova too:
```{r}
bfAnova()
```

<br> 
This finding also aligns with a recent study assessing whether engaging in a cached vs parametric (mental rotation) strategy affects implicit learning captured by strategy exclusion trials. Wang & Taylor (2026) found that strategy type does not change how much implicit learning happens, but it affects the generalization across space with algorithmic strategy users (more variability in aiming) showing more implicit generalization.y 


