---
title: "Explicit Timecourse"
output: html_notebook
editor_options:
  markdown:
    wrap: 72
---

# Overview

When participants experience a cursor rotation while making a reach to a
target, they must compensate by deviating their hand movement in order
to get a cursor straight to its target. While the strategies participants 
implement to achieve this are thought to emerge gradually, recent data
in our lab has shown that strategies might instead develop as a stepwise
process (D'Amario et al.)

In this experiment, we aimed to identify the timecourse of explicit
adaptation. Participants learned to counter a rotation and were
instructed to report their aiming strategy by changing the direction of
an arrow.

Rotation sizes introduced were 20, 30, 40, 50, and 60 degrees

<br>

# Data

We will be exploring: <br>1. The proportion of learners within each
perturbation condition. <br>2. The proportion of aiming strategies as a
function of rotation size. <br>3.The Emergence of distinct patterns across strategy-users.

# Setup

Get necessary packages:

##renv
```{r}

if (!requireNamespace("renv", quietly = TRUE)) install.packages("renv")
library(renv)

renv::restore(prompt = FALSE)

```



```{r}
library('remotes')
ip <- installed.packages()
if ('Reach' %in% ip[,'Package']) {
  if (ip[which(ip[,'Package'] == 'Reach'),'Version'] < "2025.02.16") {
    remotes::install_github('thartbm/Reach')
  }
} else {
  remotes::install_github('thartbm/Reach')
}
```



Source project code:

```{r}
# load sources for this project
source("R/data.R") 
source("R/learnerCount.R") 
source("R/strategyCount.R") 
source("R/aiming/AimDevPlot.R") 
source("R/aiming/AimHistograms.R") 

source("R/Hierarchical Clustering/FeatureSetup.R") 
source("R/Hierarchical Clustering/hierarchical.R")
source("R/Hierarchical Clustering/exparticipants.R") 
source("R/Hierarchical Clustering/glm.R")

source("R/Reach Data/ReachDevPlot.R")
source("R/Reach Data/nocursorReaches.R")
source("R/statistics.R") 
```

Download Files
```{r}
getData()
```

## Analysis

### Learning

Did all participants learn to counter the perturbation at least 50% of
the time?

```{r}
LearnerCSV()
```

This is determined by ensuring the baseline-corrected reachadeviation_deg median of the
final 16 rotated trials is greater than 50% of the rotation magnitude
received.

Then we can filter out non-learners


We can plot participant reach deviations to show the magnitude of
learning per rotation group

```{r}
plotREACH()

```

### Aiming

[Aiming analysis script](newcode.R)

Did all learners have a strategy (ie, deviate their aim)?

```{r}
strategySummary()
```

We can use a 95% CI approach to determine strategy-users. A strategy
will be defined to: 1) Within the last 8 trials of the rotation phase,
each CI bound must be above 0 degrees 2) The lower bound must be above 5
degrees to account for motor noise

For further analysis, we can create a csv file for only strategy users.

```{r}
Strategyfile()
strategy_data <- read.csv("data/strategy_only_participants.csv")
```

#### Statistics




**Does aim magnitude vary across rotation sizes?**

```{r}
aimAOV()
```


**Does reach magnitude vary across rotation sizes and strategy user?**
<br> - Two way ANOVA comparing rotation and strategy

```{r}
twoAnova()
```
```{r}
postHoc()
```


#### Visualization

**Aiming Behaviour**
```{r}
plotAIM()
```


### Are there 'patterns' in strategy development among participants?

Inspired by D'Amario et al 2024 which first showed aiming patterns different from 
an inverted exponential decay function, we noticed that not all participants develop
strategies solely as a step or as an exponential process. We note different patterns as well.
Tsay et al. 2024 initially suggested distinct classifications of
strategies. We can therefore conduct a kMeans cluster analysis to identify patterns
in our own data to determine if phenotypes other than gradual and stepwise exist.

**1) Using scaled Features derived from proposed patterns**
Features capture standard deviation (baseline normalized), trial-trial changes, linearity (R^2), smoothness, directional sign flips (learning length normalized), jump_ratio, jump dominance.

```{r}
getFeatures()
```

Determine which features vary with rotation
```{r}
kruskalFeatures()
```
```{r}
plotBox()
```


**2) Run a PCA on participant z-score features to capture aim heterogeneity**
```{r}
loadingTable()
```


**2) Determine which components are most informative**
We can create a plot that assesses cumulative variance explained across PC components. Here, we can calculate the variance of each component and divide by the variance sum. Looks like the first 3 components make the largest contribution to the total variance (~80%) with steeper slopes.
```{r}
plotScree()
```

**3) Visualize the PCA space**
```{r}
plot_pca_triplot()
```



**4) How does the PCA space vary with rotation**
```{r}
plotRot()
```

60-degree group is far right on PC1 axis, 20 degree group is far left. We can run a linear regression to determine if differences are meaningful
```{r}
lmRot()
```
Since PC1 is only significant, plot linear relation:
```{r}
lmPlot()
```


**5) Run an HCA on most informative components**

Compare clusters to visually predicted clusters
```{r}
library(dendextend)
dendogram()
```

```{r}
pcaPlot()
```

Note we identify these clusters as partions along a continuum. Because the PCA space is not very separated, this supports the idea that strategy learnin exists along a spectrum. Identified clusters may represent main regions along this spectrum, but we acknowledge the existence of intermediate strategy patterns.

**5) Do identified cluster dsitributions differ across rotation groups?**
```{r}
stackedPlot()
```

*Run GLMs*
Alternative model = likelihood of strategy * rotation
Null Model = likelihood of strategy
AICnull - AIcalt = >+, then alternative model is the better fit

likelihood of strategy development (multinomial GLM) 
```{r}
NonAIC()
```

likelihood of cluster distribution (multinomial GLM)
```{r}
make_MUmodel()
```

since deltaAIC is over 2, we can see which exact cluster differs using binmial GLM for each cluster
```{r}
clusters_to_test <- setdiff(unique(proportions_combined$cluster_label), "Non-strategy")
results <- do.call(rbind, lapply(clusters_to_test, make_model))
results
```


### No cursor reaches

After 120 rotated trials, participants undergo 24 trials of no-cursor reaches. We can use these trials to assess aftereffects and whether it changes across strategy/non-strategy-users and strategy phenotype. 

Here, we take the first 8 no cursor trials post rotation to account for all targets. We also baseline normalize from the final nocursor reach trials before the rotation. We should not expect differences in aftereffects (Modchalingam et al., 2019)

####1) Do aftereffects differ across rotation groups?
```{r}
plotRotAfter()
```
```{r}
rotAfterStats()
```


####2) Do aftereffects differ across strategy vs non strategy?
```{r}
stratAfterPlot()
```

```{r}
stratAfterStats()
```


####3) Do aftereffects differ across clusters?
```{r}
clusterAfterPlot()
```

```{r}
clusterAfterStats()
```




