---
title: "Explicit Timecourse"
output: html_notebook
editor_options:
  markdown:
    wrap: 72
---

# Overview

When participants experience a cursor rotation while making a reach to a
target, they must compensate by deviating their hand movement in order
to get a cursor straight to its target.

In this experiment, participants learned to counter a rotation and were
instructed to report their aiming strategy by changing the direction of
an arrow.

Rotation sizes introduced were 20, 30, 40, 50, and 60 degrees

<br>

# Data

We will be exploring: <br>1. The proportion of learners within each
perturbation condition. <br>2. The proportion of aiming strategies as a
function of rotation size. <br>3. The onset of an aiming strategy as a
function of rotation size. <br>4. The magnitude of the aiming strategy
as a function of rotation size.

# Setup

Get necessary packages:

##renv
```{r}

if (!requireNamespace("renv", quietly = TRUE)) install.packages("renv")
library(renv)

renv::restore(prompt = FALSE)

```



```{r}
library('remotes')
ip <- installed.packages()
if ('Reach' %in% ip[,'Package']) {
  if (ip[which(ip[,'Package'] == 'Reach'),'Version'] < "2025.02.16") {
    remotes::install_github('thartbm/Reach')
  }
} else {
  remotes::install_github('thartbm/Reach')
}
```



Source project code:

```{r}
# load sources for this project
source("R/data.R") 
source("R/newcode.R") 
source("R/AimDevPlot.R") 
source("R/Clustering/learningAlg.R") 
source("R/Clustering/KmeansXG.R") 
source("R/Clustering/clusterBehaviours.R")
source("R/Reach Data/ReachDevPlot.R")
source("R/Reach Data/nocursorReaches.R")
source("R/newstats.R") 
```

Download Files
```{r}
getData()
```

## Analysis

### Learning

Did all participants learn to counter the perturbation at least 50% of
the time?

```{r}
LearnerCSV()
```

This is determined by ensuring the baseline-corrected reachadeviation_deg median of the
final 16 rotated trials is greater than 50% of the rotation magnitude
received.

Then we can filter out non-learners


We can plot participant reach deviations to show the magnitude of
learning per rotation group

```{r}
plotREACH()

```

### Aiming

[Aiming analysis script](newcode.R)

Did all learners have a strategy (ie, deviate their aim)?

```{r}
strategySummary()
```

We can use a 95% CI approach to determine strategy-users. A strategy
will be defined to: 1) Within the last 8 trials of the rotation phase,
each CI bound must be above 0 degrees 2) The lower bound must be above 5
degrees to account for motor noise

For further analysis, we can create a csv file for only strategy users.

```{r}
Strategyfile()
strat_data <- read.csv("data/strategy_only_participants.csv")
```

#### Statistics

[Aiming stats script](newstats.R)

1)  Does final aligned and first rotated differ from \~0

```{r}
zeroT()


```

Here we took the mean final 8 aligned and mean final 8 rotated per
participant and ran 5 paired T-tests per rotation group. All tests
reveal significance.

2)  Is likelihood of strategy development predicted by rotation size?

```{r}
logAnalysis()
```

The odds of using an explicit strategy are higher in the 40°, 50°,
and 60° rotation groups compared to the 20° group.

3)  Does aim magnitude vary across rotation sizes?

```{r}
aimAOV()
```

Taking the mean from the last 16 trials, One way anova shows an effect
of rotation size. Post-hoc tukey shows that the 60 degree group produces
a higher aim strategy than all other groups.

4)  Does aim variance vary across rotation sizes?

```{r}
aimvarAOV()
```

Taking the sd from the last 16 trials, one way anova shows an effect of
rotation size. Post-hoc tukey shows that participants in the 60°
rotation group show significantly greater variability in their final
performance than participants in the 20°, 30°, and 40° groups.

5)  Is there an effect of rotation size for the timepoint of strategy-onset?

```{r}
startTrialAOV()
```

#### Visualization

**Aiming Behaviour**

```{r}
plotAIM()
```


### Are there 'patterns' in strategy development among participants?

Inspired by Tsay et al. 2024 which suggested distinct classifications of
strategies. We can thus implement a similar cluster analysis to
determine if we can observe similar patterns through three methods

**1) Predicting individual participant learning phases**

- First, we need to define a 'learning phase' finding the point of strategy onset to  when a strategy becomes stable,       using a rolling window  analysis approach, 
- Trial start: when a participants aim deviation exceeds the motor noise thresholf of 5 degrees.
- Trial end: when the SD of a participants aim deviation closely matches the SD of the final 16 aiming trials and 
  remains within an SD of 1.9.

```{r}
xgRun()
```
We can also visually assign these timepoints and compare these classifcations to the XgBoost model's predictions.

```{r}
plotStart()
```
```{r}
plotEnd()
```



**2) Run a PCA participant features to capture aim heterogeneity**

FEATURES:
1) SD of learning phase
2) Length of learning phase
3) Absolute trial to trial changes
4) Number of clockwise aims
5) Highest aim jump
6) Highest aim jump / mean of jump magnitudes
7) Number of sign flips (erraticness)
8) Smoothness in aiming - normalized sum of absolute trial-to-trial changes, providing a measure of local variability in the learning trajectory
9) linearity - using the R² of a linear regression of aim over trials, capturing the extent to which behavior followed a consistent linear trend

Run PCA to determine most informative components:
```{r}
extract <- getFeaturesFromModel() 
  
model_df <- xgRun()
strategy_data <- read.csv("data/strategy_only_participants.csv")

features_df <- extract$features_df        
k_input     <- extract$kmeans_input     
k_scaled <- scale(k_input)

# -------------------------------
# PCA 
pca <- prcomp(k_scaled, center = TRUE, scale. = FALSE)
pca
```
HIGHEST FEATURE LOADINGS:
- PC1 - length duration & eratticness (num of sign flips)
- PC2 - variance & jump magnitude (absolute difference, SD, max jump)
- PC3 & 4- gradualness (smoothness & linearity)


**2) Run an unsupervised kMeans clustering analysis on four most informative components**

Compare clusters to visually predicted clusters
```{r}
confMatrix()
```

We can see which centres closely match participants' behavioural features by plotting the two most informative components
```{r}
plotComponents()
```
Thus, we have found three clusters or 'learning phenotypes that describe strategy heterogeneity. Theres some overlap.. these particpants can be showing both!

Is cluster predicted by rotation size?

```{r}
rotationEffect() 
```

####Investigating the overlap

#### What do group averages pr cluster look like? Do we still find an inverted exponential decay function?



### No cursor reaches
####1) Do aftereffects differ across non-strategy and strategy users?

```{r}
getWashoutStrategy()
```
```{r}
washoutStats()
```

####2) Do aftereffects differ across strategy types?
```{r}
getWashoutCluster()
```
```{r}
washoutClusterStats()
```




